Carryover effects in free recall reveal how prior experiences influence memories of new experiences

### 2023-01-13 paper/main.tex
finished writing up individual difference results (up through fig. 8).   - note: i moved the "fingerprint trajectories" figure (now fig. 6) earlier in the text, since i thought it fit better in the "group results" descriptions

When we closely examined the sequences of words participants recalled in early
order manipulated lists (Fig.~\ref{fig:recall-dynamics}A, top panel), we
noticed several differences from the dynamics of participants' recalls of
randomly ordered lists (Fig.~\dynamicsRandom). One striking difference is that
participants in the category condition (dark purple curves,
Fig.~\ref{fig:recall-dynamics}) most often initiated recall with the
fourth-from-last item (\textit{Recall initiation}, top left panel), whereas
participants who recalled randomly ordered lists tended to initiate recall with
either the first or last list items (Fig.~\dynamicsRandom, top left panel). We
hypothesized that the participants might be ``clumping'' their recalls into
groups of items that shared category labels. Indeed, when we compared the
positions of feature changes in the study sequence
(Fig.~\ref{fig:recall-dynamics}B; see \textit{Identifying event boundaries})
with the positions of items participants recalled first, we noticed a striking
correspondence in both semantic conditions. Specifically, on category-ordered
lists, the category labels changed every four items on average (dark purple
peaks in Fig.~\ref{fig:recall-dynamics}B), and participants also seemed to
display an increased tendency (relative to other order manipulation and random
conditions) to initiate recall of category-ordered lists with items whose study
positions were integer multiples of four. Similarly, for size-ordered lists,
the size labels changed every eight items on average (light purple peaks in
Fig.~\ref{fig:recall-dynamics}B), and participants also seemed to display an
icnreased tendancy to initiate recall of size-ordered lists with items whose
study positions were integer multiples of eight. A second striking difference
is that participants in the category condition exhibited a much steeper lag-CRP
(Fig.~\ref{fig:recall-dynamics}A, top middle panel) than participants in other
conditions. (This is another expression of participants' increased tendencies
to temporally cluster their recalls on category-ordered lists, as we reported
above.) Taken together, these order-specific idiosyncracies suggest a
hierarchical set of influences on participants' memories. At longer timescales,
``event boundaries'' (to use the term loosely) can be induced across lists by
adding or removing irrelevant visual features. At shorter timescales, ``event
boundaries'' can be induced across items (within a single list) by adjusting
how item features change throughout the list.

@@

@@

When we examined the data at the level of individual participants
(Fig.~\ref{fig:clustering-scatterplots}), a clearer story emerged.

@@

% figure: fingerprint trajectories
\begin{figure}[tp] \centering
    \includegraphics[width=\textwidth]{figures/fingerprint_trajectories}

% figure: carryover effects: 
%  - clustering early vs. clustering late
%  - clustering carryover vs. accuracy difference
%  - clustering carryover vs. temporal clustering differences
\begin{figure}[tp] \centering
    \includegraphics[width=\textwidth]{figures/clustering_carryover}

>>>>

When we closely examined the sequences of words participants recalled in early
order manipulated lists (Fig.~\ref{fig:recall-dynamics}A, top panel), we
noticed several differences from the dynamics of participants' recalls of
randomly ordered lists (Figs.~\dynamicsRandom,~\recallInit). One striking difference is that
participants in the category condition (dark purple curves,
Fig.~\ref{fig:recall-dynamics}) most often initiated recall with the
fourth-from-last item (\textit{Recall initiation}, top left panel), whereas
participants who recalled randomly ordered lists tended to initiate recall with
either the first or last list items (Fig.~\dynamicsRandom, top left panel). We
hypothesized that the participants might be ``clumping'' their recalls into
groups of items that shared category labels. Indeed, when we compared the
positions of feature changes in the study sequence
(Fig.~\ref{fig:recall-dynamics}B; see \textit{Identifying event boundaries})
with the positions of items participants recalled first, we noticed a striking
correspondence in both semantic conditions. Specifically, on category-ordered
lists, the category labels changed every four items on average (dark purple
peaks in Fig.~\ref{fig:recall-dynamics}B), and participants also seemed to
display an increased tendency (relative to other order manipulation and random
conditions) to initiate recall of category-ordered lists with items whose study
positions were integer multiples of four. Similarly, for size-ordered lists,
the size labels changed every eight items on average (light purple peaks in
Fig.~\ref{fig:recall-dynamics}B), and participants also seemed to display an
icnreased tendancy to initiate recall of size-ordered lists with items whose
study positions were integer multiples of eight. A second striking difference
is that participants in the category condition exhibited a much steeper lag-CRP
(Fig.~\ref{fig:recall-dynamics}A, top middle panel) than participants in other
conditions. (This is another expression of participants' increased tendencies
to temporally cluster their recalls on category-ordered lists, as we reported
above.) Taken together, these order-specific idiosyncracies suggest a
hierarchical set of influences on participants' memories. At longer timescales,
``event boundaries'' (to use the term loosely) can be induced across lists by
adding or removing irrelevant visual features. At shorter timescales, ``event
boundaries'' can be induced across items (within a single list) by adjusting
how item features change throughout the list.

@@

We also looked for more subtle group-level patterns. For example, perhaps
sorting early lists by one feature dimension could affect how participants
cluster \textit{other} features (on early and/or late lists) as well. We
defined participants' \textit{memory fingerprints} as the set of temporal and
feature clustering scores. A participant's memory fingerprint describes how
they tend to retrieve memories of the studied items, perhaps searching through
several feature spaces (or along several representational dimensions). To gain
insights into the dynamics of how participants' clustering scores tended to
change over time, we computed the average (across participants) fingerprint
from each list, from each order manipulation condition
(Fig.~\ref{fig:fingerprint-trajectories}). We projected these fingerprints into
a two-dimensional space to help visualize the dynamics (top panels; see
\textit{Computing low-dimensional embeddings of memory fingerprints}). We found
that participants' average fingerprints tended to remain relatively stable on
early lists, and exhibited a ``jump'' to another stable state on later lists.
The sizes of these jumps varied somewhat across conditions (the Euclidean
distances between fingerprints in their original high dimensional spaces are
displayed in the bottom panels). We also averaged the fingerprints across early
and late lists, respectively, for each condition
(Fig.~\ref{fig:fingerprint-trajectories}B). We found that participants'
fingerprints on early lists seem to be influenced by the order manipulations on
those lists (see the locations of the circles in
Fig.~\ref{fig:fingerprint-trajectories}B). There also seemed to be some
consistency across different features within a broader type. For example, both
semantic feature conditions (category and size; purple markers) diverge in a
similar direction from the group; both lexicographic feature conditions (length
and first letter; yellow markers) diverge in a similar direction; and both
visiual conditions (color and location; green) also diverge in a similar
direction. But on late lists, participants' fingerprints seem to return to a
common state that is roughly shared across conditions (i.e., the stars in that
panel are clumped together).

\begin{figure}[tp] \centering
    \includegraphics[width=\textwidth]{figures/fingerprint_trajectories}

When we examined the data at the level of individual participants
(Figs.~\ref{fig:clustering-scatterplots} and \ref{fig:clustering-carryover}), a
clearer story emerged. Within each order manipulation condition, participants
exhibited a range of feature clustering scores, on both early and late lists
(Fig.~\ref{fig:clustering-scatterplots}A, B). Across every order manipulation
condition, participants who exhibited stronger feature clustering (for their
condition's manipulated feature) recalled more words. This trend held overall
across conditions and participants (early: $r(179) = 0.537, p < 0.001$; late: $
r(179) = 0.492, p = 0.000$) as well as for each condition individually for
early ($r$s $\geq 0.386$, all $p$s $\leq 0.035$) and late ($r$s $\geq 0.462$,
all $p$s $\leq 0.010$) lists. We found no evidence of a condition-level trend;
for example the conditions where participants tended to show stronger
clustering scores were not correlated with the conditions where participants
remembered more words (early: $r(4) = 0.526, p = 0.284$; late: $r(4) = -0.257,
p = 0.623$; see insets of panels A and B). We observed carryover associations
between feature clustering and recall performance
(Fig.~\ref{fig:clustering-scatterplots}C, D). Participants who showed stronger
feature clustering on early lists tended to recall more items on late lists
(across conditions: $r(179) = 0.492, p < 0.001$; all conditions individually:
$r$s $\geq 0.462$, all $p$s $\leq 0.010$). Participants who recalled more items
on early lists also tended to show stronger feature clustering on late lists
(across conditions: $r(179) = 0.280, p < 0.001$; all non-visual conditions:
$r$s $\geq 0.445$, all $p$s $\leq 0.014$; color: $r(29) = 0.298, p = 0.103$;
location: $r(28) = 0.354, p = 0.055$). Neither of these effects showed
condition-level trends (early feature clustering versus late recall
probability: $r(4) = -0.299, p = 0.565$; early recall probability versus late
feature clustering: $r(4) = 0.400, p = 0.432$). We also looked for associations
between feature clustering and temporal clustering. Across every order
manipulation condition, participants who exhibited stronger feature clustering
also exhibited stronger temporal clustering. For early lists
(Fig.~\ref{fig:clusteringfig:clustering-scatterplots}E), this trend held
overall ($r(179) = 0.924, p < 0.001$), for each condition individually (all
$r$s $\geq 0.822$, all $p$s $< 0.001$), and across conditions ($r(4) = 0.964, p
= 0.002$). For late lists
(Fig.~\ref{fig:clusteringfig:clustering-scatterplots}F), the results were more
variable (overall: $r(179) = 0.348, p = 0.000$; all non-visual conditions: $r$s
$\geq 0.382$, all $p$s $\leq 0.037$; color: $r(29) = 0.453, p = 0.011$;
location: $ r(28) = 0.190, p = 0.314$; across-conditions: $r(4) = -0.036, p =
0.945$). While less robust than the carryover associations between feature
clustering and recall performance, we also observed some carryover associations
between feature clustering and temporal clustering
(Fig.~\ref{fig:clustering-scatterplots}G, H). Participants who
showed stronger feature clustering on early lists trended towards showing
stronger temporal clustering on later lists (overall: $r(179) = 0.301, p <
0.001$; for individual conditions: all $r$s $\geq 0.297$, all $p$s $\leq
0.111$; across conditions: $ r(4) = 0.107, p = 0.840$). And participants who
showed stronger temporal clustering on early lists trended towards showing
stronger feature clustering on later lists (overall: $r(179) = 0.579, p <
0.001$; all non-visual conditions: $r$s $\geq 0.323$, all $p$s $\leq 0.082$;
visual conditions: $r$s $\geq 0.089$, all $p$s $\leq 0.632$; across conditions:
$r(4) = 0.916, p = 0.010$). Taken together, the results displayed in
Figure~\ref{fig:clustering-scatterplots} show that participants who were more
sensitive to the order manipulations (i.e., participants who showed stronger
feature clustering for their condition's feature on early lists) remembered
more words and showed stronger temporal clustering. These associations also
appeared to carry over across lists, even when the items on later lists were
presented in a random order.

@@

If participants show different sensitivities to order manipulations, how do
their behaviors carry over to later lists? We found that participants who
showed strong feature clustering on early lists often tended to show strong
feature clustering on late lists (Fig.~\ref{fig:clustering-carryover}A; overall
across participants and conditions: $r(179) = 0.592, p < 0.001$; non-visual
feature conditions: all $r$s $\geq 0.350$, all $p$s $\leq 0.058$; color: $r(29)
= -0.071, p = 0.704$; location: $r(28) = 0.032, p = 0.868$; across conditions:
$r(4) = 0.934, p = 0.006$). Although participants tended to show weaker feature
clustering on late lists (Fig.~\ref{fig:fingerprint-trajectories}) on
\textit{average}, the associations between early and late lists for individual
participants suggests that some influence of early order manipulations may
linger on late lists. We found that participants who exhibited larger carryover
in feature clustering (i.e., continued to show strong feature clustering on
late lists) for the semantic order manipulations (but not other manipulations)
also tended to show a larger improvement in recall
(Fig.~\ref{fig:clustering-carryover}B; overall: $r(179) = 0.378, p < 0.001$;
category: $r(28) = 0.419, p = 0.021$; size: $r(28) = 0.737, p < 0.001$;
non-semantic conditions: all $r$s $\leq 0.252$, all $p$s $\geq 0.179$; across
conditions: $r(4) = 0.773, p = 0.072$) on late lists, relative to early lists.
Participants who exhibited larger carryover in feature clustering also tended
to show stronger temporal clustering on late lists (relative to early lists)
for all but the category condition (Fig.~\ref{fig:clustering-carryover}C;
overall: $r(179) = 0.434, p < 0.001$; category: $r(28) = 0.229, p = 0.223$; all
non-category conditions: all $r$s $\geq 0.448$, all $p$s $\leq 0.012$; across
conditions: $r(4) = 0.598, p = 0.210$). We suggest two potential
interpretations of these findings. First, it is possible that some participants
are more ``maliable'' or ``adaptable'' with respect to how they organize
incoming information. When presented with list of items sorted along
\textit{any} feature dimension, they will simply adopt that feature as a
dominant dimension for organizing those items and subsequent (randomly ordered)
items. This flexibility in memory organization might afford such participants a
memory advantage, explaining their strong recall performance. An alternative
interpretation is that each participant comes into our study with a
``preferred'' way of organizing incoming information. If they happen to be
assigned to an order manipulation condition that matches their preferences,
then they will appear to be ``sensitive'' to the order manipulation and also
exhibit a high degree of carryover in feature clustering from early to late
lists. These participants might demonstrate strong recall performance not
because of their inherantly superior memory abilities, but rather because the
specific condition they were assigned to happened to be especially easy for
them, given their pre-experimental tendancies.

\begin{figure}[tp] \centering
    \includegraphics[width=\textwidth]{figures/clustering_carryover}

@@

% figure: fingerprint trajectories

% figure: carryover effects: 
%  - clustering early vs. clustering late
%  - clustering carryover vs. accuracy difference
%  - clustering carryover vs. temporal clustering differences
